# `mypackage` `README.Rmd`

This `README.Rmd` in `mypackage` should contain a description of the package, instructions on how to install it and an example.

**`README.Rmd`**

Your script should contain:

````
---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}`r ''`
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# mypackage

<!-- badges: start -->
<!-- badges: end -->

The goal of mypackage is to print a personalised greeting from me!

## Installation

You can install the development version of mypackage from GitHub with:

```{r}`r ''`
devtools::install_github("annakrystalli/mypackage")
```

## Example

This is a basic example which shows you how to print a generic greeting:

```{r example}`r ''`
library(mypackage)
hello()
```

This is a basic example which shows you how to print a personalised greeting:
```{r}`r ''`
hello("Lucy Elen")
```

````